---
title: "Time series forcasting methods: Moving average"

author: ["Nikhil Chandra Sarkar, PhD"]
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Let's get started with installation of package manager package called **pacman**

```{r}
 install.packages('packman')

```

## Load the **pacman** package using the library() function:

```{r}

library(pacman)

```

## We can use p_load() to install the remaining packages we will need for the rest of this time series data analytics

```{r}

p_load('tidyverse',      # data manipulation and visualization
       'ggplot2',        # data visualization
       'lubridate',      # working with date and time data
       'fpp2',           # working with time series data
       'zoo')            # working with time series data

```

## Here I use US economic time series data for time series analytics 

economics is a data frame with 574 rows and 6 variables.
 
# Variables name and description.
 
 date: Month of data collection.
 pce: personal consumption expenditures, in billions of dollars.
 pop: total population, in thousands.
 psavert: personal saving rate.
 uempmed: median duration of unemployment, in weeks.
 unemploy: number of unemployed in thousands.

```{r}

data <- economics # import data from ggplot2
data

```


# Get a summary of the data to help locate any potential data quality issues

```{r}

skim(data)

```

## Time series graph for personal saving rate

```{r}

ggplot(data = data, aes(x = date, y = unemploy)) +
      geom_point(color='dark blue', size=0.6) +
      geom_line(color = 'red') +
      #geom_line(color = "goldenrod") +
      theme_light() +
     coord_cartesian(xlim = c(date("1965-01-01"), date("2016-01-01")), ylim = c(2000, 20000)) +
      xlab('Month of data collection') +
      ylab('Number of unemployed in thousands')

```

## Centered Moving Averages

The most straightforward time series data analytic method is a simple moving average. For this method, we choose a number of neighborhood points and average them to estimate the trend. When calculating a simple moving average, it is beneficial to use an odd number of points so that the calculation is symmetric.

## Here first, I compute the 13 month moving average values and add this data back to the data frame.

```{r}
unemployed <- data %>%
  select(date, unemploy) %>%
  mutate(unemploy_ma13 = rollmean(unemploy, k = 13, fill = NA))

```

## Now we can go ahead and plot these values and compare the actual data to the different moving average smoothers.

```{r}

unemployed %>%
  gather(metric, value, unemploy:unemploy_ma13) %>%
  ggplot(aes(date, value, color = metric)) +
  geom_line() +
  coord_cartesian(xlim = c(date("1965-01-01"), date("2016-01-01")), ylim = c(2000, 20000)) +
      theme_light() +
      xlab('Date') +
      ylab('Number of unemployed in thousands')
  
```

## We can see this by zooming into the 2000-2015 time range:

```{r}

unemployed %>%
  gather(metric, value, unemploy:unemploy_ma13) %>%
  ggplot(aes(date, value, color = metric)) +
  geom_line() +
  coord_cartesian(xlim = c(date("2000-01-01"), date("2010-01-01")), ylim = c(2000, 20000)) +
      theme_light() +
      xlab('Date') +
      ylab('Number of unemployed in thousands')

```

## Here, I compute the 13 and 25 month moving average values and add this data back to the data frame.

```{r}
unemployed <- data %>%
  select(date, unemploy) %>%
  mutate(unemploy_ma13 = rollmean(unemploy, k = 13, fill = NA),
         unemploy_ma25 = rollmean(unemploy, k = 25, fill = NA)
        )

```


## Now we can go ahead and plot these values and compare the actual data to the different moving average smoothers.

```{r}

unemployed %>%
  gather(metric, value, unemploy:unemploy_ma25) %>%
  ggplot(aes(date, value, color = metric)) +
  geom_line() +
  coord_cartesian(xlim = c(date("1965-01-01"), date("2016-01-01")), ylim = c(2000, 20000)) +
      theme_light() +
      xlab('Date') +
      ylab('Number of unemployed in thousands')
  
```
## To understand how these different moving averages compare we can compute the MSE and MAPE. Both of these error rates will increase as you choose a larger k to average over.

```{r}

unemployed %>%
  gather(metric, value, unemploy_ma13:unemploy_ma25) %>%
  group_by(metric) %>%
  summarise(MSE = mean((unemploy - value)^2, na.rm = TRUE),
            MAPE = mean(abs((unemploy - value)/unemploy), na.rm = TRUE))

```

## Trailing Moving Average for Forecasting

```{r}
unemployed_tma <- data %>%
  select(date, unemploy) %>%
  mutate(unemploy_tma = rollmean(unemploy, k = 24, fill = NA, align = "right"))

tail(unemployed_tma, 15)

```


## We can visualize how the 24-month trailing moving average predicts future number of unemployed in thousands with the following plot.

```{r}
unemployed_tma %>%
  gather(metric, value, -date) %>%
  ggplot(aes(date, value, color = metric)) +
  geom_line()+
  xlab('Date') +
  ylab('Number of unemployed in thousands')

```
# Here, I compute the MSE and MAPE errors

```{r}
unemployed_tma %>%
  gather(metric, value, unemploy_tma) %>%
  group_by(metric) %>%
  summarise(MSE = mean((unemploy - value)^2, na.rm = TRUE),
            MAPE = mean(abs((unemploy - value)/unemploy), na.rm = TRUE))

```

## Moving Averages of Moving Averages


```{r}

economics %>% 
  mutate(ma4 = ma(unemploy, order = 4, centre = TRUE)) %>%
  head()
  
```


## To compare this moving average to a regular moving average we can plot the two outputs:


```{r}
# compute 2 and 2x4 moving averages
 economics %>%
  mutate(ma2 = rollmean(unemploy, k = 2, fill = NA),
         ma2x4 = ma(unemploy, order = 4, centre = TRUE)) %>%
  gather(ma, value, ma2:ma2x4) %>%
  ggplot(aes(x = date)) + 
  geom_point(aes(y = unemploy)) +
  geom_line(aes(y = value, color = ma))+
  xlab('Date') +
  ylab('Number of unemployed in thousands')

```

